---
title: "P8105 Homework 3"
output: github_document
---

Name: Xi Peng   UNI: xp2213    Date: 10.14.2024

```{r setup, include=FALSE}

library(tidyverse)
library(p8105.datasets)
library(dplyr)
library(patchwork)
library(ggridges)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

# Question 1. Exploration of "NY NOAA" dataset

## Section 1: Data cleaning

```{r, message=FALSE}

data("ny_noaa")

weather_dat = ny_noaa |> 
  janitor::clean_names() |> 
  separate(date, into = c('year', 'month', 'day'), remove = TRUE) |>
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin)
  )

snow_comm = weather_dat |> 
  group_by(snow,id) |> 
  summarise(count = n()) |> 
  arrange(desc(count))
  
```

In the original "ny_noaa" datasets, there are `r nrow(ny_noaa)` observations and there are `r ncol(ny_noaa)` variables included, which are: `r names(ny_noaa)`. In this section, the "date" variable was separated into year, month, and day. Temperature ("tman" and "tmin"), precipitation ("prcp"), and snowfall("snow") were tranfered into reasonable units.

By checking the "snow" variable for snowfall, the result shows that `0 cm` and `NA` are the most commonly observed values. The likely reason is that for most of the year, these regions do not experience snowfall, especially during warmer months. Additionally, some regions included in the dataset may not experience snow at all throughout the year due to their warmer climates or geographical location. These places are unlikely to have a snow season, and as a result, `0 cm` is consistently recorded for snowfall in these areas. The presence of `NA` values indicates missing data, which could be due to weather stations not recording snowfall on certain days or data collection issues.


## Section 2: Comparison of mean maximum temperature in January and July across weather stations over the years

```{r}

weather_jan_jul = weather_dat |> 
  group_by(id, year, month) |> 
  mutate(
    month = as.numeric(month),
    year = as.numeric(year)
         ) |>
  filter(month %in% c(1, 7)) |>
  summarise(mean_tmax = mean(tmax, na.rm = TRUE, color = id))
 
weather_jan_jul_ggplot = weather_jan_jul |> 
ggplot(aes(x = year, y = mean_tmax, group = id)) +
  geom_point() + geom_path() +
  facet_grid(~month) +
  scale_x_continuous(breaks = seq(1980, 2010, by = 10)) +
  labs(
    title = "Mean Max Temperature in January and July by Stations",
       x = "Year", 
       y = "Mean Max Temperature (°C)") 

weather_jan_jul_ggplot

```

I checked the dataset and recognized there are `r length(unique(pull(weather_dat,id)))` stations. One issue that needed attention is the large number of stations in the dataset. Including all the station labels would result in the scatterplot not being displayed properly because of the excessive number of stations in the legend.

According to the scatterplot, there are clear seasonal differences between January and July temperatures. The temperatures are generally colder in January, and warmer in July across stations and years. Due to the large number of stations included in this dataset, without restricting the analysis to a specific time range or a particular set of stations, I can only observe general trends. The observable structure shows that temperatures in January typically range from -10 °C to 10 °C, where temperatures in July typically ranger from 20 °C to 32.5 °C. Comparing 1981 to 2010, there appears to be a slight increase in temperatures for both January and July, suggesting a possible warming trend over time.There's considerable variability both between stations and from year to year, illustrating geographical and annual climate differences. There are several noticeable outlines. For example, in January 1982 and 1996, some stations recorded average maximum temperatures below -10 °C, reflecting extremely cold winters. Also, one station in July 1988 recorded unusually low average temperature, far below the general trend. These anomalies could indicate extreme weather events or errors in data recording.


## Section 3: Analysis of temperature extremes and distribution of snowfall patterns

```{r, message=FALSE, warning=FALSE}

tmax_vs_tmin_panel = weather_dat |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex() +
  labs(
    title = "Temperature Comparison: tmax vs tmin",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)"
  ) 

snowfall_distri_panel = weather_dat |> 
  filter(snow < 100, snow > 0) |> 
  ggplot(aes(x = snow, y = as.factor(year))) +
  geom_density_ridges() +
  labs(
    subtitle = "Distribution of snowfall values by year",
    x = "Snow",
    y = "as.factor(year)"
  ) 

tmax_vs_tmin_panel + snowfall_distri_panel

```



# Question 2. Analysis of accelerometer data from the NHANES study

## Section 1: Dataset import and organization

```{r}

nhanes_covar_df =
    read_csv("Data/nhanes_covar.csv", na = c("NA", "", ".", " "), skip = 4) |> 
    janitor::clean_names() |> 
  drop_na() |> 
  mutate( education = as.character(education), sex = as.character(sex)
  ) |> 
    mutate(
      education = case_match(
        education,
        "1" ~ "Less than high school",
        "2" ~ "High school equivalent",
        "3" ~ "More than high school"),
      sex = case_match(
        sex,
        "1" ~ "Male",
        "2" ~ "Female"
      ))
    
nhanes_accel_df =
    read_csv("Data/nhanes_accel.csv", na = c("NA", "", ".", " ")) |> 
  janitor::clean_names()

Merged_nhanes_dat = nhanes_covar_df |> 
  inner_join(nhanes_accel_df, by = "seqn") |> 
  filter(age > 21) |> 
  mutate(
    sex = factor(sex, levels = c("Male", "Female")),
    Education = factor(education, levels = c(
      "Less than high school",
      "High school equivalent",
      "More than high school"
    )))

```

In this section, the datasets were cleaned and organized as per the requirements, and a merged dataset was created. In this dataset, there are `r nrow(Merged_nhanes_dat)` observations and `r ncol(Merged_nhanes_dat)` variables.

## Section 2: Gender distribution and age trends across education categories

```{r, message=FALSE, fig.width = 12}

Sex_in_education_cate = Merged_nhanes_dat |> 
  group_by(Education, sex) |> 
  summarise(count = n()) |> 
  pivot_wider(
    names_from = sex,
    values_from = count
  )

knitr::kable(Sex_in_education_cate, caption = "Number of Men and Women in Each Education Category")

```

According to the table, the number of males and females education levels of "Less than high school" and "More than high school" are approximately similar, inidacting a balanced distribution between genders in these catgories. For education level of "High school equivalent", males outnumber females by eleven, showing that men are more likely to obtain a high school equivalent degree compared to famles in this sample population.

```{r}

Age_distri_in_Edc_plot = Merged_nhanes_dat |> 
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.4, adjust = 0.5) +
  facet_grid(~Education) +
  labs(
    title = "Age Distribution by Education and Gender",
    x = "Age",
    y = "Density"
  ) 

Age_distri_in_Edc_plot

```

According to the density plot, three education levels are showing different age distributions between genders. For the "Less than high school" education category, males have a peak between age 40 and 50, while females show a more even distribution across age. Both genders display a left-skewed distribution, with higher densities in older age groups. For "High school equivalent" category, males display a relatively uneven distribution of age with various peaks across ages, while females show a more left-skewed distribution with a peak between age 70 and 80. Lastly, for the "More than high school" category, both genders demonstrate a right-skewed distribution. Females have a prominent peak at age of 30, and males also show less pronounced peak around that age.

## Section 3: Impact of age, gender, and education on total daily activity

```{r}

Total_daily_activity = Merged_nhanes_dat |> 
  group_by(seqn, age, sex, Education) |> 
  summarise(total_activity = sum(c_across(starts_with("min")), na.rm = TRUE)) |> 
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(~Education) +
  labs(
    title = "Total Daily Activity vs. Age by Education Level and Gender",
    x = "Age",
    y = "Total Activities"
  )
 
Total_daily_activity  

```

For the first category, "Less than high school," as age rises, overall activity gradually decreases for both males and females. Females show more variability in activity levels at later ages, whereas males tend to start with higher overall activity at earlier ages. However, by the time they are 60 years old and beyond, the gap between males and females narrows considerably. For participants with "High school equivalent" degrees, the overall activity of females peaks more prominently at the age of 30, after which it gradually decreases. In contrast, males experience a more extensive peak in their activity levels between the ages of 30 and 50, after which they drastically decrease. In this educational group, females are often more active than males. In the last category, "More than high school," activity levels for both sexes are comparatively constant from early years until about age 60, at which point they noticeably decline. While in younger age groups, women seem to be more active than men, the pattern gradually reverses in middle age and then levels off as both sexes get older.


## Section 4: 24Hr activity patterns by education level and gender

```{r}

hourly_activity = Merged_nhanes_dat |>
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minutes",
    values_to = "activity",
    names_prefix = "min"
  ) |> 
  mutate(minutes = as.numeric(minutes)) |> 
  ggplot(aes(x = minutes, y = activity, color = sex)) +
  geom_smooth(se = FALSE) +
  facet_grid(~education) +
  labs(
    title = "24Hr Activity Patterns by Education Level and Gender",
    x = "Minute of the Day",
    y = "Activity (MIMS)"
  )

hourly_activity

```

According to the graph, the "Less than high school" group demonstrates the highest activity levels during the day compared to the other education categories. Females tend to maintain slightly higher activity levels than males across all education categories, although the differences between genders are not very pronounced. Additionally, most people tend to engage in higher activity between approximately 500 minutes and 1250 minutes, with lower activity during the early morning and late evening.Afterwards, the activity level dropped.

# Problem 3  NYC Citi Bike System Data Exploration

## Section 1: Dataset import and clean

```{r}

NYCiti_Jan_2020 =
    read_csv("Data/Jan 2020 Citi.csv", na = c("NA", "", ".", " ")) |> 
  janitor::clean_names() |> 
  mutate(
    year = 2020, month = "Jan"
  )

NYCiti_Jan_2024 =
    read_csv("Data/Jan 2024 Citi.csv", na = c("NA", "", ".", " ")) |> 
  janitor::clean_names() |> 
  mutate(
    year = 2024, month = "Jan"
  )

NYCiti_July_2020 =
    read_csv("Data/July 2020 Citi.csv", na = c("NA", "", ".", " ")) |> 
  janitor::clean_names() |> 
  mutate(
    year = 2020, month = "July"
  )

NYCiti_July_2024 =
    read_csv("Data/July 2024 Citi.csv", na = c("NA", "", ".", " ")) |> 
  janitor::clean_names() |> 
  mutate(
    year = 2024, month = "July"
  )

Total_NYCiti = 
  bind_rows(NYCiti_Jan_2020, NYCiti_Jan_2024, NYCiti_July_2020, NYCiti_July_2024) |> 
  janitor::clean_names()

Total_NYCiti_summary = Total_NYCiti |> 
  group_by(year, month, member_casual) |> 
  summarise(total_riders = n()) |> 
  pivot_wider(
    names_from = member_casual,
    values_from = total_riders
  )

knitr::kable(Total_NYCiti_summary)

```


## Section 2: The five most popular starting stations for July 2024

```{r}

Five_most_pop_stationsJu2024 = NYCiti_July_2024 |> 
  group_by(start_station_name) |> 
  summarise(number_of_rides = n()) |> 
  arrange(desc(number_of_rides)) |> 
  head(5)

knitr::kable(Five_most_pop_stationsJu2024,
             col.names = c("Start Station Name", "Total Number of Rides"),
             caption = "The Top Five Starting Stations for July 2024"
             )

```

## Section 3: The effects of day of the week, month, and year on median ride duration

```{r}

effects_on_median_duration = Total_NYCiti |> 
  group_by(year, month, weekdays) |> 
  summarise(median_ride_duration = median(duration)) |> 
  mutate(
    weekdays, factor(weekdays, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
    ) |> 
  ggplot(aes(x = weekdays, y = median_ride_duration, group = year, color = factor(year))) +
  geom_line() +
  geom_point() +
  facet_grid(~month) +
  labs(
    title = "The Effects of Day of the Week, Month, and Year on Median Ride Duration",
    x = "Day of the week",
    y = "Median Ride Duration (mins)"
  )
  
effects_on_median_duration

```

According to the plots, ride durations in July are typically longer than in January, with this pattern being particularly evident in the 2020 data set. In July 2020, the median ride duration reaches its peak on the weekends, whereas in January 2020, it remains relatively stable throughout the week. By comparison, the 2024 data set exhibits less discrepancy in the median ride duration across the seven-day period in both months. The values observed in July are consistently higher than those recorded in January for both years. This indicates that ride durations are subject to seasonal variations, with July displaying longer rides, potentially due to favorable weather conditions. Furthermore, there is a noticeable discrepancy between 2020 and 2024, with the median ride durations consistently lower in 2024 across both months and days of the week.

## Section 4: The impact of month, membership status, and bike type on the distribution of ride duration in 2024

```{r}

distri_duration2024 = Total_NYCiti |> 
  filter(year == "2024") |> 
  ggplot(aes(x = month, y = duration, fill = member_casual)) + 
  geom_violin(trim = TRUE) + 
  facet_grid(~rideable_type) + 
  labs(
    title = "Impact of Month, Membership Status, and Bike Type on Ride Duration in 2024", 
    x = "Month", 
    y = "Ride Duration (mins)", 
    fill = "Membership Status"
  )

distri_duration2024

```

According to the plots, electric bikes consistently show shorter ride durations compared to classic bikes, suggesting they might be favored for quicker trips. Membership status significantly impacts ride duration, with casual riders taking longer trips than members across all categories. Seasonal differences are evident, with July rides generally lasting longer than January rides, particularly for casual users on classic bikes. This seasonal effect is less pronounced for electric bikes, indicating their utility year-round. Casual riders exhibit more variability in ride duration, especially with classic bikes, while members show more consistent usage patterns.