---
title: "P8105 Homework 3"
output: github_document
date: "2024-10-12"
Name: "Xi Peng"   
UNI: "xp2213"
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)
library(dplyr)
```

#Question 1 Exploration of "NY NOAA" dataset

##Section 1 Data cleaning

```{r, message=FALSE}

data("ny_noaa")

weather_dat = ny_noaa |> 
  janitor::clean_names() |> 
  separate(date, c('year', 'month', 'day'), sep ='-', remove = TRUE) |>
  mutate(
    tmax = as.numeric(tmax) / 10,
    tmin = as.numeric(tmin) / 10,
    prcp = prcp / 10,
    snow = snow / 10,
  )

snow_comm = weather_dat |> 
  group_by(snow,id) |> 
  summarise(count = n()) |> 
  arrange(desc(count))
  
```

In the original "ny_noaa" datasets, there are `r nrow(ny_noaa)` observations and there are `r ncol(ny_noaa)` variables included, which are: `r names(ny_noaa)`. In this section, the "date" variable was separated into year, month, and day. Temperature ("tman" and "tmin"), precipitation ("prcp"), and snowfall("snow") were tranfered into reasonable units.

By checking the "snow" variable for snowfall, the result shows that `0 cm` and `NA` are the most commonly observed values. The likely reason is that for most of the year, these regions do not experience snowfall, especially during warmer months. Additionally, some regions included in the dataset may not experience snow at all throughout the year due to their warmer climates or geographical location. These places are unlikely to have a snow season, and as a result, `0 cm` is consistently recorded for snowfall in these areas. The presence of `NA` values indicates missing data, which could be due to weather stations not recording snowfall on certain days or data collection issues.


